---
title: "Hayman Fire Recovery"
author: "Mikaela Richardson"
date: "02/07/2022"
output: html_document

---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir='..')

library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)
library(ggpubr)

# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways


```

# Hayman Fire Recovery Assignment

For this assignment, we worked with data wrangling and becoming familiarized with project organization. In doing so, we analyzed data regarding vegetation recovery of the 2002 Hayman Fire.

## Reading in Data and Stacking it

```{r dataread}
# Read in individual data files
ndmi <- read_csv("C:/Users/mikae/Documents/School/CSU/Spring_2022/Data_Science/Intro/Final Project/FinalProjectBookdown/data/assignment2/hayman_ndmi.csv") %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')


ndsi <- read_csv("C:/Users/mikae/Documents/School/CSU/Spring_2022/Data_Science/Intro/Final Project/FinalProjectBookdown/data/assignment2/hayman_ndsi.csv") %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv("C:/Users/mikae/Documents/School/CSU/Spring_2022/Data_Science/Intro/Final Project/FinalProjectBookdown/data/assignment2/hayman_ndvi.csv")%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))


```

## Assignment

### Question 1 

What is the correlation between NDVI and NDMI? - here I want you to
convert the full_long dataset in to a wide dataset using the 
function "spread" and then make a plot that shows the correlation as a
function of if the site was burned or not (x axis should be ndmi)
You should exclude winter months and focus on summer months
 
```{r, message=F, fig.cap = "Scatterplot of NDVI versus NDMI for the Hayman Fire burn scar. There is a positive correlation between NDVI and NDMI with a value of 0.96. As moisture index increases so does vegetation index. In general, the plot indicates that for unburned sites the majority of them have high NDVI values as compared to the burned sites. For both types of sites, the plots indicate positive relationships."}
# converting long data to wide
full_wide <- spread(data=full_long,key='data',value='value') %>%
  filter_if(is.numeric,all_vars(!is.na(.))) %>%
  mutate(month = month(DateTime),
         year = year(DateTime))

#filter out winter months
summer_only <- filter(full_wide,month %in% c(6,7,8,9))

#plotting ndmi vs ndvi for summer months
ggplot(summer_only,aes(x=ndmi,y=ndvi,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8)) +
  labs(y="NDVI", x="NDMI")

cor(summer_only$ndmi, summer_only$ndvi)
```


### Question 2 

2) What is the correlation between average NDSI (normalized
 snow index) for January - April and average NDVI for June-August?
In other words, does the previous year's snow cover influence vegetation
 growth for the following summer?


```{r, message=F, fig.cap = "Average NDSI for January - April versus average NDVI for June - August in the Hayman Fire burn scar. Based on the correlation value of 0.18 there is a positive relationship meaning the previous year's snow cover does influence vegetation growth for the following summer. However, based on the low value it does not indicate a strong correlation between the two as compared to NDVI and NDMI in Question 1."}
# summarizing the data by year
ndvi_annual <- full_wide[c("DateTime", "ndvi", "month", "year", "site")] %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(site,year) %>%
  summarize(ndvi_annual=mean(ndvi))

ndsi_annual <- full_wide[c("DateTime", "ndsi", "month", "year", "site")] %>%
  filter(month %in% c(1,2,3,4)) %>%
  group_by(site,year) %>%
  summarize(ndsi_annual=mean(ndsi))

combined <- inner_join(ndsi_annual, ndvi_annual)

cor(combined$ndsi_annual, combined$ndvi_annual)

ggplot(combined,aes(x=ndsi_annual,y=ndvi_annual,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8)) +
  labs(y="NDVI", x="NDSI")
```


### Question 3

How is the snow effect from question 2 different between pre- and post-burn
and burned and unburned? 

```{r, message=F}
#pre burned correlation
ndvi_pre <- full_wide[c("DateTime", "ndvi", "month", "year", "site")] %>%
  filter(month %in% c(6,7,8)) %>%
  filter(year %in% c(1984:2001)) %>%
  group_by(site,year) %>%
  summarize(ndvi_pre=mean(ndvi))

ndsi_pre <- full_wide[c("DateTime", "ndsi", "month", "year", "site")] %>%
  filter(month %in% c(1,2,3,4)) %>%
  filter(year %in% c(1984:2001)) %>%
  group_by(site,year) %>%
  summarize(ndsi_pre=mean(ndsi))

combined_pre <- inner_join(ndsi_pre, ndvi_pre)

cor(combined_pre$ndsi_pre, combined_pre$ndvi_pre)

a <- ggplot(combined_pre,aes(x=ndsi_pre,y=ndvi_pre,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8)) +
  labs(y="NDVI", x="NDSI", title = "a. Pre-burn")

```

```{r,message=F}
#post burned correlation
ndvi_post <- full_wide[c("DateTime", "ndvi", "month", "year", "site")] %>%
  filter(month %in% c(6,7,8)) %>%
  filter(year %in% c(2003:2019)) %>%
  group_by(site,year) %>%
  summarize(ndvi_post=mean(ndvi))

ndsi_post <- full_wide[c("DateTime", "ndsi", "month", "year", "site")] %>%
  filter(month %in% c(1,2,3,4)) %>%
  filter(year %in% c(2003:2019)) %>%
  group_by(site,year) %>%
  summarize(ndsi_post=mean(ndsi))

combined_post <- inner_join(ndsi_post, ndvi_post)

cor(combined_post$ndsi_post, combined_post$ndvi_post)

b <- ggplot(combined_post,aes(x=ndsi_post,y=ndvi_post,color=site)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8)) +
  labs(y="NDVI", x="NDSI", title = "b. Post-burn")
```

```{r, message=F}
#burned sites correlation
ndvi_burn <- full_wide[c("DateTime", "ndvi", "month", "year", "site")] %>%
  filter(month %in% c(6,7,8)) %>%
  filter(site %in% "burned") %>%
  group_by(year) %>%
  summarize(ndvi_burn=mean(ndvi))

ndsi_burn <- full_wide[c("DateTime", "ndsi", "month", "year", "site")] %>%
  filter(month %in% c(1,2,3,4)) %>%
  filter(site %in% "burned") %>%
  group_by(site,year) %>%
  summarize(ndsi_burn=mean(ndsi))

combined_burned <- inner_join(ndsi_burn, ndvi_burn)

cor(combined_burned$ndsi_burn, combined_burned$ndvi_burn)

c <- ggplot(combined_burned,aes(x=ndsi_burn,y=ndvi_burn)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.8,0.8)) +
  labs(y="NDVI", x="NDSI", title = "c. Burned sites")
```

```{r, message=F, fig.cap = "a. NDVI vs. NDSI in pre-burned areas. b. NVDI vs. NDSI for post-burned areas. c. NDVI vs. NDSI for all burned sites. d. NDVI vs. NDSI for all unburned sites."}
#unburned sites correlation
ndvi_unburned <- full_wide[c("DateTime", "ndvi", "month", "year", "site")] %>%
  filter(month %in% c(6,7,8)) %>%
  filter(site %in% "unburned") %>%
  group_by(year) %>%
  summarize(ndvi_unburned=mean(ndvi))

ndsi_unburned <- full_wide[c("DateTime", "ndsi", "month", "year", "site")] %>%
  filter(month %in% c(1,2,3,4)) %>%
  filter(site %in% "unburned") %>%
  group_by(site,year) %>%
  summarize(ndsi_unburned=mean(ndsi))

combined_unburned <- inner_join(ndsi_unburned, ndvi_unburned)

cor(combined_unburned$ndsi_unburned, combined_unburned$ndvi_unburned)

d <- ggplot(combined_unburned,aes(x=ndsi_unburned,y=ndvi_unburned)) + 
  geom_point() + 
  theme_few() + 
  scale_color_few() + 
  labs(y="NDVI", x="NDSI", title = "d. Unburned sites")

ggarrange(a,b,c,d, nrow = 2, ncol = 2)
```
Pre-burned correlation: 0.091
Post-burned correlation:0.24
Burned sites correlation: 0.087
Unburned sites correlation: -0.031

There are positive correlations for NDVI and NDSI between pre- and post-burn dates and burned sites, and a negative correlation for unburned sites. Based on the correlation values and plots, the correlations are not noticeably strong.

### Question 5

What month is the snowiest on average?

```{r, message=F, fig.cap = "Histogram looking at average NDSI for every month for the Hayman Fire burn area. Based on NDSI, the snowiest month on average is January."}
ndsi_mean <- full_wide[c("ndsi", "month")] %>%
  group_by(month) %>%
  summarize(ndsi_mean=mean(ndsi))

max(ndsi_mean$ndsi_mean)

ggplot(ndsi_mean, aes(x=month,y=ndsi_mean))+
  geom_bar(stat="identity") +
  labs(y="Average NDSI", x= "Month")
```


