
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(rvest)
library(tidyverse)
library(lubridate)
library(readxl)

```

# Snow Data Assignment 

This assignment focused on webscraping, iterations and functions. Data was scraped from the Snow Studies website and for loops were used to read in the data. Analysis looked at temperatures over time for the Swamp Angel and Senator Beck Study Plots.

## Simple web scraping

R can read html using either rvest, xml, or xml2 packages. Here we are going to navigate to the Center for Snow and Avalance Studies  [Website](https://snowstudies.org/archived-data/) and read a table in. This table contains links to data we want to programatically download for three sites. We don't know much about these sites, but they contain incredibly rich snow, temperature, and precip data. 


## Reading an html 

### Extract CSV links from webpage

```{r}
site_url <- 'https://snowstudies.org/archived-data/'
#Read the web url
webpage <- read_html(site_url)
#See if we can extract tables and get the data that way
tables <- webpage %>%
  html_nodes('table') %>%
  magrittr::extract2(3) %>%
  html_table(fill = TRUE)
#That didn't work, so let's try a different approach
#Extract only weblinks and then the URLs!
links <- webpage %>%
  html_nodes('a') %>%
  .[grepl('24hr',.)] %>%
  html_attr('href')
```

## Data Download

### Download data in a for loop

```{r}
#Grab only the name of the file by splitting out on forward slashes
splits <- str_split_fixed(links,'/',8)
#Keep only the 8th column
dataset <- splits[,8] 
#generate a file list for where the data goes
file_names <- paste0('data/assignment3/',dataset)
for(i in 1:3){
  download.file(links[i],destfile=file_names[i])
}
downloaded <- file.exists(file_names)
evaluate <- !all(downloaded)
```


### Download data in a map

```{r}
#Map version of the same for loop (downloading 3 files)
if(evaluate == T){
  map2(links[1:3],file_names[1:3],download.file)
}else{print('data already downloaded')}
```

## Data read-in 

### Read in just the snow data as a loop

```{r}
#Pattern matching to only keep certain files
snow_files <- file_names %>%
  .[!grepl('SG_24',.)] %>%
  .[!grepl('PTSP',.)]
#empty_data <- list()
# snow_data <- for(i in 1:length(snow_files)){
#   empty_data[[i]] <- read_csv(snow_files[i]) %>%
#     select(Year,DOY,Sno_Height_M)
# }
#snow_data_full <- do.call('rbind',empty_data)
#summary(snow_data_full)
```


### Read in the data as a map function

```{r}
our_snow_reader <- function(file){
  name = str_split_fixed(file,'/',2)[,2] %>%
    gsub('_24hr.csv','',.)
  df <- read_csv(file) %>%
    select(Year,DOY,Sno_Height_M) %>%
    mutate(site = name)
}
snow_data_full <- map_dfr(snow_files,our_snow_reader)
summary(snow_data_full)
```


### Plot snow data

```{r}
snow_yearly <- snow_data_full %>%
  group_by(Year,site) %>%
  summarize(mean_height = mean(Sno_Height_M,na.rm=T))
ggplot(snow_yearly,aes(x=Year,y=mean_height,color=site)) + 
  geom_point() +
  ggthemes::theme_few() + 
  ggthemes::scale_color_few()
```

## Assignment

### Question 1. Extract the meteorological data URLs. Here we want you to use the `rvest` package to get the URLs for the `SASP forcing` and `SBSP_forcing` meteorological datasets.
```{r}
site_url <- 'https://snowstudies.org/archived-data/'

webpage <- read_html(site_url)

links <- webpage %>%
  html_nodes('a') %>%
  .[grepl('forcing',.)] %>%
  html_attr('href')
```


### Question 2. Download the meteorological data. Use the `download_file` and `str_split_fixed` commands to download the data and save it in your data folder. You can use a for loop or a map function. 
```{r}
splits <- str_split_fixed(links,'/',8)

dataset <- splits[,8] 

file_names <- paste0('data/assignment3/',dataset)

#for(i in 1:3){
  #download.file(links[i],destfile=file_names[i])
#}

#downloaded <- file.exists(file_names)

#evaluate <- !all(downloaded)
```


### Question 3. Write a custom function to read in the data and append a site column to the data. 

```{r, message = F}
# this code grabs the variable names from the metadata pdf file
library(pdftools)
headers <- pdf_text('https://snowstudies.org/wp-content/uploads/2022/02/Serially-Complete-Metadata-text08.pdf') %>%
  readr::read_lines(.) %>%
  trimws(.) %>%
  str_split_fixed(.,'\\.',2) %>%
  .[,2] %>%
  .[1:26] %>%
  str_trim(side = "left")

our_forcing_reader <- function(file_names){
  name = str_split_fixed(file_names,'/',2)[,2]
  name2 = str_split_fixed(file_names,'_',4)[,2]
  df <- read.delim(file_names, header = FALSE, sep ="", col.names = headers, skip = 4) %>%
    select(1:14) %>%
    mutate(site = name2)
}
```

### Question 4. Use the `map` function to read in both meteorological files. Display a summary of your tibble.
```{r}
forcing_data_full <- map_dfr(file_names,our_forcing_reader)

summary(forcing_data_full)
```


### Question 5. Make a line plot of mean temp by year by site (using the `air temp [K]` variable). Is there anything suspicious in the plot? Adjust your filtering if needed.

The year 2003 had lower average temperatures compared to the following years which could be attributed to missing data resulting in lower averages. I decided to filter out this year so it was more representative.
```{r, message = F, fig.cap = "A time series of average monthly temperatures (K) at the Swamp Angel Study Plot and the Senator Beck Study Plot."}
temp_yearly <- forcing_data_full %>%
  filter(year > 2003) %>%
  group_by(year,site) %>%
  summarize(mean_temp = mean(air.temp..K.,na.rm=T))

ggplot(temp_yearly,aes(x=year,y=mean_temp,color=site)) + 
  geom_line() +
  ggthemes::theme_few() + 
  ggthemes::scale_color_few() +
  labs(x="Year",
       y="Average Temperature (K)",
       color = "Site") +
  scale_color_manual(labels = c("Swamp Angel", "Senator Beck"), values = c("green", "orange")) +
  theme(legend.position = "bottom")
```


### Question 6. Write a function that makes line plots of monthly average temperature at each site for a given year. Use a for loop to make these plots for 2005 to 2010. Are monthly average temperatures at the Senator Beck Study Plot ever warmer than the Swamp Angel Study Plot?
Hint: https://ggplot2.tidyverse.org/reference/print.ggplot.html


```{r, message = F}
lineplotter <- function(forcing_data_full,year){
  temp_monthly <- forcing_data_full %>%
    group_by(month, year, site) %>%
    summarize(mean_temp = mean(air.temp..K.)) %>%
        filter (yr == year)
  
  plots <- ggplot(temp_monthly, aes (x = month, y = mean_temp, color = site)) +
    geom_line()+
    labs(x= "Month",
         y = "Average Temperature (K)",
         title = temp_monthly$year,
         color = "Site") +
    scale_color_manual(labels = c("Swamp Angel", "Senator Beck"), values = c("green", "orange"))
  
  print(plots)
}

years <- c(2005, 2006, 2007, 2008, 2009, 2010)

for (yr in years){
  lineplotter(forcing_data_full, year)
}
```
Figure 2. Monthly average temperatures for 2005-2010 for Senator Beck and Swamp Angel study plots. It appears that the Senator Beck Study Plot is never warmer than the Swamp Angel Study Plot. The largest differences are during the summer months and the smallest occurring during the fall months.

