
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
```


# LAGOS Analysis Part 1

This assignment was focused on spatial operations and analysis in R and was split in two parts. For the first part, we downloaded data from Lagos and converted to spatial data using the sf function. The spatial data was then used to analyze lake distribution and size in Minnestoa, Illionis, and Iowa.

## Loading in data


### First download and then specifically grab the locus (or site lat longs)

```{r data-read}
# #Lagos download script
#LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())


#Load in lagos
lagos <- lagosne_load()

#Grab the lake centroid info
lake_centers <- lagos$locus
```



### Convert to spatial data
```{r}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>%
  st_transform(2163)

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)

```


### Subset to only Minnesota

```{r}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```



## Assignment


### Question 1. Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}
iowa <- states %>%
  filter(name == "Iowa") %>%
  st_transform(2163)

illinois <- states %>%
  filter(name == "Illinois") %>%
  st_transform(2163)

iail <- rbind(iowa,illinois)

  mapview(iail)
  
```



### Question 2. Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

Minnesota has ~29,000 lakes while Illinois and Iowa combined only have ~16,500 lakes. This is understandable as Minnesota is known as the land of many lakes.
```{r}
iail_lakes <- spatial_lakes[iail,]
```


### Question 3. What is the distribution of lake size in Iowa vs. Minnesota?

- Here I want to see a histogram plot with lake size on x-axis and frequency on y axis (check out geom_histogram)

```{r, fig.cap = "The distribution of lake sizes in Iowa vs. Minnesota. Both states share a similar distribution of lakes, with the majority being smaller than 1 ha. However, based on the histograms, it is demonstrated how many more lakes Minnesota contains based on the frequencies of the lake sizes."}
#combining Iowa and Minnesota
iamn <- rbind(iowa,minnesota)

#subsetting lakes
iamn_lakes <- spatial_lakes %>%
  .[iamn,] %>%
  st_join(iamn) 

ggplot()+
  geom_histogram(filter(iamn_lakes, name == 'Iowa'), mapping = aes(lake_area_ha), bins = 20, color = "red", fill = "orange")+
  scale_x_log10()+
  labs(x = 'Lake Area (ha)', y = 'Frequency') +
  geom_histogram(filter(iamn_lakes, name == "Minnesota"), mapping = aes(lake_area_ha),bins = 20, color = "blue", fill = "light green")+
  scale_x_log10()+
  labs(x = 'Lake Area (ha)', y = 'Frequency', title = 'Iowa vs Minnesota Lakes Size Distribution')+
    facet_wrap(~name)
```


Histogram code used from: http://www.sthda.com/english/wiki/ggplot2-histogram-plot-quick-start-guide-r-software-and-data-visualization and WR418 assignment, Accessed: 2/23/2022

### Question 4. Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r}
iail_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```

### Question 5. What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

Other data sources we could use to understand reservoirs and natural lakes vary in size would be remote sensing imagery. Landsat has been used in several studies looking at lake sizes, as well as how they are changing over time. With this continuous data source, great insight could be provided to visualize the variation of water bodies in these 3 states.

