
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

# Weather Data Analysis Assignment

This assignment focused on looking at linear and quadratic regressions for temperature trends and corn and soy yields.

## Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/assignment5/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```

```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & doy <= 59 | doy >= 335 & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r}
# set our API key with NASS
nassqs_auth(key = "34702FFD-D70D-3C61-9A25-83550E7443D6")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")
```


```{r yield download, results = FALSE}
# download
cornyieldsall <- nassqs_yields(params)
```


```{r}
cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)
```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r, fig.cap = 'Time series of corn yield in Winneshiek County. It is demonstrated that there is a clear, positive trend of corn yield over time. Additionally, the linear model has a p-value <0.05, providing evidence of a significant time trend for corn yields.'}
win_corn <- cornyields %>%
  filter(county_name=="WINNESHIEK")%>%
  select(yield,year)

model1 <- lm(win_corn$yield ~ win_corn$year)
summary(model1)

ggplot(win_corn, aes(x=year,y=yield))+
  geom_point()+
  geom_smooth(method = 'lm', formula = y ~ x, se = T)+
  labs(y="Corn Yield ",x="Year", title="Time Series of Corn Yield in Winneshiek County")
```


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 
```{r, fig.cap = 'Plot of corn yield versus time for Winneshiek County with a quadratic time trend line. Based on the model, the trend is linear and not curved meaning that there is no evidence for a slowing yield growth.'}
win_corn$yearsq <- win_corn$year^2

win_quad <- lm(yield ~ year + yearsq, win_corn)
summary(win_quad)
win_corn$fitted <- win_quad$fitted.values

ggplot(win_corn) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "Year", y = "Corn Yield", title = "Quadratic Time Trend of Corn Yields in Winneshiek County")
```


### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r, fig.cap = 'Plot of corn yield versus summer temperature in Winneshiek County with a quadratic trend model. Based on the model, there is negative relationship between temperature and corn yields with time. Looking at the quadratic versus linear model is helpful in identifying this trend.'}
win_corn_temp <- inner_join(win_corn, winnesummer, by = "year")

win_corn_temp$tempsq <- win_corn_temp$meantmax^2

#linear model
win_temp_model <- lm(yield ~ meantmax, win_corn_temp)
summary (win_temp_model)
win_corn_temp$fitted_linear <- win_temp_model$fitted.values

#quadratic model
win_temp_quad <- lm(yield ~ meantmax + tempsq, win_corn_temp)
summary(win_temp_quad)
win_corn_temp$fitted <- win_temp_quad$fitted.values

ggplot(win_corn_temp)+
  geom_point(mapping = aes(y=yield, x = meantmax)) +
  geom_line(mapping = aes(y= fitted, x = meantmax)) +
  labs(x = "Average Max Temperature (C)", y = "Corn Yield", title = "Quadratic Trend of Temperature and Corn Yields in Winneshiek County")

```

### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r, fig.cap = 'Scatterplot of summer temperatures and corn yields in 2018 for all counties with a linear model. Based on the linear model summary, the p-value is >0.05, providing no evidence of a relationship between temperature and corn yields across all counties in 2018.'}
#Yield and temp for 2018
corn_18 <- cornyields %>%
  filter(year=="2018")%>% 
  group_by(county_name) %>% 
  unique() %>% 
  filter(!is.na(county_ansi))

summer_temp_18 <- tmaxdf %>%
  group_by(countyfp) %>%
  filter(year == 2018) %>% 
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax)) %>% 
  rename(county_ansi = "countyfp")

summer_temp_18$county_ansi <- as.numeric(as.character(summer_temp_18$county_ansi))

#Join temp and yield data
data_2018 <- inner_join(corn_18, summer_temp_18, by = "county_ansi")

#Linear model
model2 <- lm(data_2018$meantmax ~ data_2018$yield)
summary(model2)

ggplot(data_2018, aes (x = yield, y = meantmax))+
  geom_point()+
  geom_smooth(method = 'lm', formula = y ~ x, se = T)+
  labs(x = "Corn Yield", y = "Average Mean Temperature (C)", title = "Temperature vs. Corn Yields for all Counties in 2018")
```


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.
```{r, fig.cap = 'The plot looks at actual yield vs fitted in relation to time, average maximum summer temperatures, and county. The plot shows that predicted yields are fairly accurately compared to the actual yields based on the point density and tightness to the blue line. The panel regression model summary has a p-value <0.05 providing evidence that there is a significant relationship between corn yield and all the additional variables.'}
#max summer temps for all counties & years
maxtempssummer <- tmaxdf %>%
  group_by(countyfp,year) %>%
  filter(doy >= 152 & doy <= 243) %>%
  summarize(meantmax = mean(tmax))%>%
  rename(county_ansi = "countyfp")

#making the county_ansi into a factor
maxtempssummer$county_ansi <- as.factor(maxtempssummer$county_ansi)
cornyields$county_ansi <- as.factor(cornyields$county_ansi)

#combined yield and summer temperature for all counties & years
county_summer_yield <- inner_join(maxtempssummer,cornyields,)%>%
  unique()

#Squaring temp
county_summer_yield$meantmaxsq <- county_summer_yield$meantmax^2

#model
lm_county_summer_yield <- lm(yield ~ meantmax + meantmaxsq + county_ansi + year, county_summer_yield)
#summary(lm_county_summer_yield)

#fitted values
county_summer_yield$fittedyield <- lm_county_summer_yield$fitted.values

#plot
ggplot(county_summer_yield, aes(x = fittedyield, y = yield)) +
  geom_point() +
  geom_abline(color = "blue", size = 1) +
  labs(x = "Actual Yield", y = "Fitted Yield", title = "Actual Corn Yield vs. Fitted Corn Yield")
```


### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.
```{r}
# parameters to query on 
params <- list(commodity_desc = "SOYBEANS", statisticcat_desc = "YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")
```


```{r, results = FALSE}
# download
soybeanyieldsall <- nassqs_yields(params)
```


```{r, fig.cap = "Time series for soybean yields in Buena Vista County. It is demonstrated that there is a clear, positive trend of soybean yield over time. Additionally, the linear model has a p-value <0.05, providing evidence of a significant time trend for soybean yields."}
soybeanyieldsall$county_ansi <- as.numeric(soybeanyieldsall$county_ansi)
soybeanyieldsall$yield <- as.numeric(soybeanyieldsall$Value)

# clean the dataset
soybeanyields <- select(soybeanyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soybeanyields <- tibble(soybeanyields)

## buena vista soy
bv_soy <- soybeanyields %>%
  filter(county_name=="BUENA VISTA")%>%
  select(yield,year)

bvmodel <- lm(bv_soy$yield ~ bv_soy$year)
summary(bvmodel)

ggplot(bv_soy, aes(x=year,y=yield))+
  geom_point()+
  geom_smooth(method = 'lm', formula = y ~ x, se = T)+
  labs(y="Soybean Yield ",x="Year", title="Time Series of Soybean Yield in Buena Vista County")

```

